Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-07 RevisedSubmission/hines_hetland_2017_gji.tex
did more

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left[u_i(p),u_i(p')\right] = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere, which ensures that $u_i$ is spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe.  To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We first consider component $i$ of a single GNSS displacement observation made at station $j$, which is located at $\vec{x}^{(j)}$, and time $t$. We describe this observation as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d^{(j)}_i(t) = &u_i(\vec{x}^{(j)},t) + \eta_i(\vec{x}^{(j)},t) + a_i^{(j)} + b_i^{(j)}t + c_i^{(j)}\sin(2 \pi t) + \\ 
               &e_i^{(j)}\cos(2 \pi t) + f_i^{(j)}\sin(4 \pi t)  + g_i^{(j)}\cos(4 \pi t), 
\end{split}
\end{align}
where $a_i^{(j)}$ is an offset that is unique for each station, $b_i^{(j)}$ is the secular velocity at $\vec{x}^{(j)}$, and the sinusoids describe seasonal deformation (using units of years for $t$). We use $\eta_i$ to represent all the non-parametric noise processes. For example, $\eta_i$ can consist of uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we only make the assumption that $\eta_i$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $b_i^{(j)}$, are spatially correlated and we could invoke a tectonic model to form a prior on $b_i^{(j)}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $b_i^{(j)}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

We move on to consider the column vector of $n$ GNSS displacement observations, $\mitbf{d}^*_i$, where the subscript indicates that we are still only considering component $i$ of displacements. The times and positions for each observation are described by the set $\mitbf{P}$. $\mitbf{d}^*_i$ can be considered a realization of the random vector $\mitbf{d}_i$, which is formed by evaluating eq. (\ref{eq:Data}) at each point in $\mitbf{P}$. To write out $\mitbf{d}_i$ explicitly, we collect the coefficients from eq. (\ref{eq:Data}) into the vector $\mitbf{m}_i$, which contains 6 elements for each of the $m$ GNSS station. We use $\mitbf{G}$ to represent the $n \times 6m$ matrix of corresponding basis functions evaluated at each point in $\mitbf{P}$. We can write $\mitbf{d}_i$ as 
\begin{equation}\label{eq:DataVec}
\mitbf{d}_i = u_i(\mitbf{P}) + \eta_i(\mitbf{P}) + \mitbf{G}\mitbf{m}_i,
\end{equation}
which is distributed as $\mathcal{N}(\mitbf{0},\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T)$, where
\begin{equation}\label{eq:Cd}
\mitbf{\Sigma} = C_{u_i}(\mitbf{P},\mitbf{P}) + C_{\eta_i}(\mitbf{P},\mitbf{P}).  
\end{equation}
It should be understood that notation such as $u_i(\mitbf{P})$ and $C_{u_i}(\mitbf{P},\mitbf{P})$ represents the column vector $[u_i(p)]_{p \in \mitbf{P}}$ and the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$, respectively.

@@

To describe our outlier detection algorithm in detail, we let $\mitbf{d}^*$ denote all $n$ GNSS displacement observations for a single component (note that we dropped the component subscripts since outliers are detected for the easting and northing components independently). Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. For notational ease, we use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). We now consider $\mitbf{d}^*$ to be a realization of the random vector
\begin{equation}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}
where $\mitbf{G}$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}), and $\mitbf{w}$ is a vector of uncorrelated noise with assumed known standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$. Our best fitting model for $\mitbf{d}^*$ is taken to be the mean of $\mitbf{G}\mitbf{m} + v(\mitbf{P})$ after it is conditioned with non-outlier observations. The residual vector is then
\begin{equation}\label{eq:Residual1}
\mitbf{r} = \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right].
\end{equation}
Following a similar procedure as above, we can write eq. (\ref{eq:Residual1}) as
\begin{equation}\label{eq:Residual}
\mitbf{r} = \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{equation}
We start the algorithm by letting $\mitbf{\Omega}$ contain all $n$ indices, and we refine $\mitbf{\Omega}$ iteratively. For each iteration, we compute $\mitbf{r}$ with eq. (\ref{eq:Residual}) and then update $\mitbf{\Omega}$ to be
\begin{equation}\label{eq:Update}
\mitbf{\Omega} = \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \mathrm{RMS}\right\},
\end{equation} 
where RMS is the root-mean-square of $\tilde{\mitbf{r}}$ and $\lambda$ is an outlier tolerance. Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$.

@@

>>>>

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{Cov}\left[u_i(p),u_i(p')\right] = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}(p,p') = C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. We also require that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere, which ensures that $u_i$ is spatially and temporally differentiable (See \citet[sec. 2.2]{Adler1981} for a formal definition of differentiation for Gaussian processes). For example, the squared exponential function is a commonly used covariance function that is sufficiently differentiable for our purposes. We can make a potential covariance function for $u_i$ by using a squared exponential for $X$ and $T$, 
\begin{equation}\label{eq:ExampleCov}
C_{u_i}\left(p,p'\right) = \phi^2\exp\left(\frac{-||\vec{x} - \vec{x}'||_2^2}{2\ell^2}\right)
                           \exp\left(\frac{-|t - t'|^2}{2\tau^2}\right),
\end{equation}
where $\ell$ and $\tau$ describe a characteristic length-scale and time-scale for realizations of $u_i$. The parameter $\phi$ controls the amplitude of realizations of $u_i$. We refer to the parameters of a covariance function, such as $\ell$, $\tau$, and $\phi$, as hyperparameters. These hyperparameters can be chosen subjectively or they can be chosen objectively with maximum likelihood techniques.

commonly used covariance function that is sufficiently differentiable for our purposes.

The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe.  To keep this section sufficiently general, we hold off on specifying $X_i$ and $T_i$ until Section \ref{sec:SignalModel}, where we estimate transient strain from slow slip events in the Pacific Northwest.

GNSS data records transient displacements as well as other physical and non-physical processes that we are not interested in. We first consider component $i$ of a single GNSS displacement observation made at station $j$, which is located at $\vec{x}^{(j)}$, and time $t^{(k)}$. We describe this observation as a realization of the random variable 
\begin{align}\label{eq:Data}
\begin{split}
d_i^{(jk)} = &u_i\left(\vec{x}^{(j)},t^{(k)}\right) + 
              \eta_i\left(\vec{x}^{(j)},t^{(k)}\right) + 
              a_i^{(j)} + b_i^{(j)}t^{(k)} + \\
             &c_i^{(j)}\sin\left(2 \pi t^{(k)}\right) +  
              e_i^{(j)}\cos\left(2 \pi t^{(k)}\right) + \\ 
             &f_i^{(j)}\sin\left(4 \pi t^{(k)}\right)  + 
              g_i^{(j)}\cos\left(4 \pi t^{(k)}\right), 
\end{split}
\end{align}
where $a_i^{(j)}$ is an offset that is unique for each station, $b_i^{(j)}$ is the secular velocity at $\vec{x}^{(j)}$, and the sinusoids describe seasonal deformation (using units of years for $t^{(k)}$). We use $\eta_i$ to represent all the non-parametric noise processes. For example, $\eta_i$ can consist of uncorrelated white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, and/or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}. For now, we only make the assumption that $\eta_i$ is a Gaussian process distributed as $\eta_i \sim \mathcal{N}(0,C_{\eta_i})$, and $C_{\eta_i}$ is a spatially and temporally dependent covariance function. We consider the coefficients in eq. (\ref{eq:Data}) to be uncorrelated random variables distributed as $\mathcal{N}(0,\kappa^2)$ in the limit as $\kappa \to \infty$ (i.e., the coefficients have diffuse priors). Of course, the secular velocities, $b_i^{(j)}$, are spatially correlated and we could invoke a tectonic model to form a prior on $b_i^{(j)}$. However, in our application to the Pacific Northwest, we will be using displacement time series which are long enough to sufficiently constrain $b_i^{(j)}$ for each station, avoiding the need to incorporate a prior. Likewise, seasonal deformation is spatially correlated \citep{Dong2002,Langbein2008}, and it may be worth exploring and exploiting such a correlation in a future study.

We move on to consider the column vector of $n$ GNSS displacement observations, $\mitbf{d}^*_i$, where the subscript indicates that we are still only considering component $i$ of displacements. The observations are made at $m$ stations, and the times and positions for each observation are described by the set $\mitbf{P}$. $\mitbf{d}^*_i$ can be considered a realization of the random vector $\mitbf{d}_i$, which is formed by evaluating eq. (\ref{eq:Data}) at each point in $\mitbf{P}$. To write out $\mitbf{d}_i$ explicitly, we let $\mitbf{G}$ be an $n \times 6m$ matrix consisting of the basis functions from eq. (\ref{eq:Data}) (i.e., the linear trends and sinusoids for each station) evaluated at each point in $\mitbf{P}$. The coefficients corresponding to each basis function are collected into the column vector $\mitbf{m}_i$. We can then write $\mitbf{d}_i$ as 
\begin{equation}\label{eq:DataVec}
\mitbf{d}_i = u_i(\mitbf{P}) + \eta_i(\mitbf{P}) + \mitbf{G}\mitbf{m}_i,
\end{equation}
which is distributed as $\mathcal{N}(\mitbf{0},\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T)$, where
\begin{equation}\label{eq:Cd}
\mitbf{\Sigma} = C_{u_i}(\mitbf{P},\mitbf{P}) + C_{\eta_i}(\mitbf{P},\mitbf{P}).  
\end{equation}
It should be understood that notation such as $u_i(\mitbf{P})$ and $C_{u_i}(\mitbf{P},\mitbf{P})$ represents the column vector $[u_i(p)]_{p \in \mitbf{P}}$ and the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$, respectively.

@@

To describe our outlier detection algorithm in detail, we let $\mitbf{d}^*$ denote all $n$ GNSS displacement observations for a single component (note that we dropped the component subscripts since outliers are detected for the easting and northing components independently). Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. For notational ease, we use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). We now consider $\mitbf{d}^*$ to be a realization of the random vector
\begin{equation}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}
where $\mitbf{G}$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}), and $\mitbf{w}$ is a vector of uncorrelated noise with assumed known standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$. Our best fitting model for $\mitbf{d}^*$ is taken to be the mean of $\mitbf{G}\mitbf{m} + v(\mitbf{P})$ after it is conditioned with non-outlier observations. The residual vector is then
\begin{equation}\label{eq:Residual1}
\mitbf{r} = \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right].
\end{equation}
Following a similar procedure as above, we can write eq. (\ref{eq:Residual1}) as
\begin{equation}\label{eq:Residual}
\mitbf{r} = \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{equation}
We start the algorithm by letting $\mitbf{\Omega}$ contain all $n$ indices, and we refine $\mitbf{\Omega}$ iteratively. For each iteration, we compute $\mitbf{r}$ with eq. (\ref{eq:Residual}) and then update $\mitbf{\Omega}$ so that it contains the indices of $\mitbf{r}$ whose weighted values are less than $\lambda$ times the weighted root mean square of $\tilde{\mitbf{r}}$, 
\begin{equation}\label{eq:Update}
\mitbf{\Omega} \leftarrow \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \sqrt{\frac{1}{|\mitbf{\Omega}|} \sum_{j \in \mitbf{\Omega}} \frac{r_j^2}{\sigma_j^2}} \right\}.
\end{equation} 
Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$.

@@

\begin{figure*}
\includegraphics{figures/outliers_new/outliers.pdf}
\caption{Detrended easting component of displacements at station SC03, which is located on Mount Olympus in Washington. The orange markers indicate outliers that were automatically detected using the algorithm from Section \ref{sec:Outlier}. The error bars show one standard deviation uncertainties. Note that outliers tend be observed in the winter, suggesting that they were caused by snow or ice.}   
\label{fig:Outliers}
\end{figure*}
